Matrix Transposition and Symmetry: a case study on Superscalar Architectures\\

### 2024-11-25 latex/paper.tex
transpose: asm implementation

Contemporary CPU architectures leverage superscalar designs to enhance performance, incorporating multiple cores and advanced Single Instruction Multiple Data (SIMD) extensions, such as Intel's Streaming SIMD Extensions (SSE) and Advanced Vector Extensions (AVX). Although SSE is not a new technology, being first introduced in the Pentium III in 1999 and subsequently expanded in later years \cite{b1}, modern algorithms leveraging SIMD instructions continue to be a topic of active research in recent literature.
For example, the FreeBSD's SIMD-enhanced libc is actively reworking the standard library's code. \cite{b2}.
In 2023, most of the string functions were reimplemented by Robert Clausecker et al on many architectures \cite{b3}, exploiting vector instructions and improving the overall performance of the standard library \cite{b4}.
Similarly, efficient multi-threaded algorithms are becoming more critical as the demand of computation increases worldwide due to AI training \cite{b5} \cite{b6}.
This case study examines the implementations and performance of two algorithms—out-of-place square matrix transposition (matTranspose) and square matrix symmetry checking (checkSymm)—that leverage superscalar architectures. Both algorithms are used pervasively in performance critical applications. For example, matrix transposition was used in NASA's Apollo 11 Guidance Computer (AGC) to convert vectors in platenary coordinate system to base reference system \cite{b7}. Additionally, matrix transposition is used by Fourier Transform in the West (FFTW) \cite{b8}, a popular library to calculate Fast Furier Transofrms, and in signal processing \cite{b9}. Furthermore, the current C++26 draft proposes basic linear algebra algorithms in the "linalg" header, including matrix transposition \cite{b10}.

In section 2, we discuss the state of the art of the two algorithms.
In section 3, we elaborate on the algorithms analyzed, discuss performance evaluation and we present a new benchmarking framework.
In section 4, we analyze experimental data and graphs.

@@

- Review current research and developments related to your project \\
- Discuss existing solutions and their limitations \\
- Identift the gap your project aims to fill \\

@@

Computing the transpose of a matrix is a well discussed problem.

Older fortran implementations were difficult due to fortran.

# a lot of In place and rectangular

Analysis of In-Situ Transposition
"Algorithms for in-place matrix transposition make use of the cyclic structure of
the transposition mapping. A clear introduction to the process was given in
1958 by Windley [6]. In 1968 Boothroyd presented ACM Algorithm 302 [1]. Then
in 1970 Laflin and Brefner presented ACM Algorithm 380 which they showed to
be significantly faster than Algorithm 302 [4]"

	[1] Transpose vector stored array.
	BOOTHROYD,Z. Algorithm 302: Transpose vector stored array. Go~r~m.A~M I0, ~ (1967),
	292-293

	anche 380 important, revised by 513

Naive algorithm does not exploit locality of reference. Recursive solutions were proposed.

## out-of-core - where I/O needs to be minimized

 The oldest context in which the spatial locality of transposition seems to have been studied is for out-of-core operation (by Alltop, 1975), where the matrix is too large to fit into main memory ("core"). 
 Reviews of out-of-core algorithms, especially as applied to parallel computing, can be found in e.g. Suh & Prasanna (2002) and Krishnamoorth et al. (2004).

 An Efficient Algorithm for
 Out-of-Core Matrix Transposition
 Jinwoo Suh, Member, IEEE, and Viktor K. Prasanna, Fellow, IEEE

 	"These efforts have focused on reducing the number
 	of I/O operations. However, in state-of-the-art architectures, memory-memory data transfer time and index computation time are also
 	significant components of the overall time. In this paper, we propose an algorithm that considers the index computation time and the
 	I/O time and reduces the overall execution time."

FFTW - MPI shit
/* Recursive "radix-r" distributed transpose, which breaks a transpose
over p processes into p/r transposes over r processes plus r
transposes over p/r processes.  If performed recursively, this
produces a total of O(p log p) messages vs. O(p^2) messages for a
direct approach.
*/

## GPUs

In-Place Transposition of Rectangular Matrices on Accelerators

An Efficient Matrix Transpose in CUDA C/C++
Feb 18, 2013
By Mark Harris

\textit{A square matrix $A$ is called symmetric if it equals its transpose} \cite{b100}.

@@

\blindtext\blindtext

@@

\blindtext\blindtext

@@

\blindtext\blindtext

@@

\section*{References}

Please number citations consecutively within brackets \cite{b1}. The 
sentence punctuation follows the bracket \cite{b2}. Refer simply to the reference 
number, as in \cite{b3}---do not use ``Ref. \cite{b3}'' or ``reference \cite{b3}'' except at 
the beginning of a sentence: ``Reference \cite{b3} was the first $\ldots$''

Number footnotes separately in superscripts. Place the actual footnote at 
the bottom of the column in which it was cited. Do not put footnotes in the 
abstract or reference list. Use letters for table footnotes.

Unless there are six authors or more give all authors' names; do not use 
``et al.''. Papers that have not been published, even if they have been 
submitted for publication, should be cited as ``unpublished'' \cite{b4}. Papers 
that have been accepted for publication should be cited as ``in press'' \cite{b5}. 
Capitalize only the first word in a paper title, except for proper nouns and 
element symbols.

For papers published in translation journals, please give the English 
citation first, followed by the original foreign-language citation \cite{b6}.

\begin{thebibliography}{00}
\bibitem{b1} Intel® 64 and IA-32 Architectures Software Developer’s Manual Combined Volumes: 1, 2A, 2B, 2C, 2D, 3A, 3B, 3C, 3D, and 4. Volume 1, Section 2.2.7 "SIMD Instructions"
\bibitem{b2} SIMD man page "https://man.freebsd.org/cgi/man.cgi?simd(7)"
\bibitem{b3} Git blame on freeBSD's source code, in "lib/libc/amd64/string", shows recent activity
\bibitem{b4} https://freebsdfoundation.org/blog/a-sneak-peek-simd-enhanced-string-functions-for-amd64/
\bibitem{b5} Amodei, D. Hernandez, D. AI and Compute, https://openai.com/blog/ai-and-compute
\bibitem{b6} Andrew J. Lohn, Micah Musse. AI and Compute How Much Longer Can Computing Power Drive Artificial Intelligence Progress? CSET Issue Brief
\bibitem{b7} Digitalized Apollo 11 guidance computer by Virtual AGC and MIT Museum. Code in "Apollo-11/Comanche055/PLANETERY\_INTERNAL\_ORIENTATION.agc" line 34; "Apollo-11/Luminary099/ATTITUDE\_MANEUVER\_ROUTINE.agc" line 441
\bibitem{b8} Matteo Frigo and Steven G. Johnson. "The design and implementation of FFTW3". The word "transpose" is mentioned 21 times
\bibitem{b9} M. Henriksson and O. Gustafsson, "Streaming Matrix Transposition on FPGAs Using Distributed Memories," 2023 IEEE Nordic Circuits and Systems Conference (NorCAS), Aalborg, Denmark, 2023, pp. 1-6, doi: 10.1109/NorCAS58970.2023.10305472.
\bibitem{b10} "https://www.open-std.org/jtc1/sc22/wg21/docs/papers/2023/p1673r12.html" Accessed in november 2024
\bibitem{b11} Sheldon Axler. Linear Algebra Done Right ourth edition
16 November 2024. Section 3C Matrices, definition 3.54
\bibitem{b12} Sheldon Axler. Linear Algebra Done Right ourth edition
16 November 2024. Section 9A Bilinear Forms and Quadratic Forms, definition 9.11

\end{thebibliography}
\vspace{12pt}

>>>>

Modern CPUs use superscalar architectures to improve performance, combining multiple cores with advanced Single Instruction Multiple Data (SIMD) technologies. Examples include Intel's Streaming SIMD Extensions (SSE) and Advanced Vector Extensions (AVX). Although SSE is not a new technology, being first introduced in the Pentium III in 1999 and subsequently expanded in later years \cite{b1}, modern algorithms leveraging SIMD instructions continue to be a topic of active research in recent literature.
For example, the FreeBSD's SIMD-enhanced libc is actively reworking the standard library's code. \cite{b2}.
In 2023, most of the string functions were reimplemented by Robert Clausecker et al on many architectures \cite{b3}, exploiting vector instructions and improving the overall performance of the standard library \cite{b4}.
Similarly, efficient multi-threaded algorithms are becoming more critical as the demand of computation increases worldwide due to AI training \cite{b5} \cite{b6}.
This case study examines the implementations and performance of two algorithms—out-of-place square matrix transposition (matTranspose) and square matrix symmetry checking (checkSymm)—in the context of a superscalar architecture. Both algorithms have been used pervasively in performance critical applications. For example, matrix transposition was used in NASA's Apollo 11 Guidance Computer (AGC) to convert vectors in platenary coordinate system to base reference system \cite{b7}. Additionally, matrix transposition is used by Fourier Transform in the West (FFTW) \cite{b8}, a popular library to calculate Fast Furier Transofrms, and in signal processing \cite{b9}. Furthermore, the current C++26 draft proposes basic linear algebra algorithms in the "linalg" header, including matrix transposition \cite{b10}.

In section 2, we discuss the state of the art of the two algorithms.
In section 3, we elaborate on the algorithms analyzed, discuss performance evaluation and we present a new benchmarking framework.
In section 4, we show experimental data and graphs that is analyzed
in section 5. The conclusion is in section 6.

@@

@@

Computing the transpose of a matrix is a well discussed computational problem. Early
work focuses on in-situ transporision, that is transposing a matrix without
relying on additional space. Implementations were often based on the cyclic
structure of the transposition mapping: values of a matrix of size $N$ can be stored
in a single contiguous array or size $N$*$N$ and indexed with $row*\#rows+ column$.
Remarcable contributions include Boothroyd's ACM Algorithm 302 (1966) using cyclic mapping \cite{b13}, Laflin and Brefner ACM Algorithm 380
adding fixed points arithmetic \cite{b14} and
a later revision of the two by Esko G. Cate and David W. Twigg in ACM Algorithm 513 \cite{b15} .
SIMD implementations for specific architectures are also widely adopted
in performance critical applications like cryptographic libraries using instructions like Intel's vpunpckldq and vpunpckhdq \cite{b16} reporting improvements up to 70\% \cite{b17}.
However, matrix operations are not efficient as vector registers access data only linearly, motivating research on new vector register file (VFS) designs such as diagonal registers \cite{b23}.
Another area of research includes out-of-core matrix transposition, that
is computing the transpose of a matrix that exceeds the size of the available in-core memory.
This problem focuses on optimizing the number of I/O operations and file layout since memory-memory data transfer is notably slower than math operations. Work in out-of-core transposition include W. 0. Alltop (1975) \cite{b18}, Suh \& Prasanna (2002) \cite{b19} and Krishnamoorth et al. (2004) \cite{b20}.
Recent work explores parallelization and the use of GPUs to offload algebra operations. Examples are I-Jui Sung et al \cite{b21} and Mark Harris \cite{b22}.

\textit{A square matrix $A$ is called symmetric if it equals its transpose} \cite{b12}.

While symmetric matrces are actively used in various areas of research as their
properties allow for several optimizations in various problems, an in depth case study of symmetry
checking on modern architectures is yet to be published.

@@

\iffalse
Remember: write
- what you did (contribution): high level overview like "hey, i want to test different vectorization ond parallelization implementations"
- how you did it (methodology): we benchmarked algorithm A ... (all the pseudocodes). Whay we mesured

\fi

Quiesti qui bellissimi e stra fighi
- uno (pseudocodice)
- due (pseudocodice)
- tre (pseudocodice)
- ...
Ho usato implicit ed esplicit parallelism con vectorization e openmp
Voglio fare questo per comparare quest'altro cos' pel meme
Benchmarkare è difficle perchè... quindi ho fatto ciò... (aplia qua)

@@

All relevant cluster info, compiler, flags, c++ standard. Test flags and data. Benchmarks metrics collected
Example:
System Description:
The experiments were conducted on a system equipped with an Intel Core i9-12900K CPU (16 cores, 24 threads, base clock 3.2 GHz), 32 GB of DDR5 RAM, and an NVIDIA RTX 3080 GPU (10 GB GDDR6X). The operating system was Ubuntu 22.04, and all code was compiled using GCC 12.2 with the -O3 optimization flag. Benchmarks were run in an isolated environment with minimal background processes to ensure consistency.

Experimental Setup:
The study benchmarked three sorting algorithms: QuickSort, MergeSort, and HeapSort. Input datasets ranged from 1,000 to 1,000,000 elements and included random, sorted, and reverse-sorted orders. Each experiment was repeated 10 times, with execution time and memory usage recorded. Performance metrics were averaged across runs to account for variability.

Mamma mia che bel framework che ho

Ho runanto tutto sul pbs con questi cores bellissimi con le ottimizzazioni diverse. Ho collezionato i dativi in un basino e l'ho grafato con quella mmerda di pitone. Ho fatto i test così perchè mi piacevano.

@@

cinquantamila grafici qui

Mamma mia ma quanto e' veloce sto asesmbly,
buu al compilatore che non ci azzecca per bene. 
Anche ciclico è nammerda, invece l'unrolling
è figo.
In generale abbiamo capito
cose ed il cluster è pazzesco e le differenze
sono notevoli, evviva i soldi.

@@

Ciao ciao

\begin{thebibliography}{00}
\bibitem{b1} Intel® 64 and IA-32 Architectures Software Developer’s Manual Combined Volumes: 1, 2A, 2B, 2C, 2D, 3A, 3B, 3C, 3D, and 4. Volume 1, Section 2.2.7 "SIMD Instructions"
\bibitem{b2} SIMD man page "https://man.freebsd.org/cgi/man.cgi?simd(7)"
\bibitem{b3} Git blame on freeBSD's source code, in "lib/libc/amd64/string", shows recent activity
\bibitem{b4} https://freebsdfoundation.org/blog/a-sneak-peek-simd-enhanced-string-functions-for-amd64/
\bibitem{b5} Amodei, D. Hernandez, D. AI and Compute, https://openai.com/blog/ai-and-compute
\bibitem{b6} Andrew J. Lohn, Micah Musse. AI and Compute How Much Longer Can Computing Power Drive Artificial Intelligence Progress? CSET Issue Brief
\bibitem{b7} Digitalized Apollo 11 guidance computer by Virtual AGC and MIT Museum. Code in "Apollo-11/Comanche055/PLANETERY\_INTERNAL\_ORIENTATION.agc" line 34; "Apollo-11/Luminary099/ATTITUDE\_MANEUVER\_ROUTINE.agc" line 441
\bibitem{b8} Matteo Frigo and Steven G. Johnson. "The design and implementation of FFTW3". The word "transpose" is mentioned 21 times
\bibitem{b9} M. Henriksson and O. Gustafsson, "Streaming Matrix Transposition on FPGAs Using Distributed Memories," 2023 IEEE Nordic Circuits and Systems Conference (NorCAS), Aalborg, Denmark, 2023, pp. 1-6, doi: 10.1109/NorCAS58970.2023.10305472.
\bibitem{b10} "https://www.open-std.org/jtc1/sc22/wg21/docs/papers/2023/p1673r12.html" Accessed in november 2024
\bibitem{b11} Sheldon Axler. Linear Algebra Done Right ourth edition
16 November 2024. Section 3C Matrices, definition 3.54
\bibitem{b12} Sheldon Axler. Linear Algebra Done Right ourth edition
16 November 2024. Section 9A Bilinear Forms and Quadratic Forms, definition 9.11
\bibitem{b13} BOOTHROYD,Z. Algorithm 302: Transpose vector stored array. Go~r~m.A~M I0, ~ (1967),
292-293
\bibitem{b14} LAFLIN, S., AND BREFNER, M.A. Algorithm 380: In-situ transposition of a rectangular
matrix, Comm. ACM IS, 5 (1970), 324-326
\bibitem{b15} ESKO G. CATE and DAVID W. TWIGG ALGORITHM 513: Analysis of In-Situ Transposition (1977)
\bibitem{b16} most of the crypto functions are written in assembly usin SIMD instructions in the
linux kernel "linux/arch/x86/crypto" for exanple transpose\_4x4 in "serpent-avx2-asm\_64.S", "sm4-aesni-avx-asm\_64.S",
"cast6-avx-x86\_64-asm\_64.S" and others
\bibitem{b17} https://github.com/torvalds/linux/commit/d34a460092d857f1616e39eed7eac6f40cea2225
\bibitem{b18} W. O. Alltop, "A Computer Algorithm for Transposing Nonsquare Matrices," in IEEE Transactions on Computers, vol. C-24, no. 10, pp. 1038-1040, Oct. 1975, doi: 10.1109/T-C.1975.224124.
\bibitem{b19} Jinwoo Suh and V. K. Prasanna, "An efficient algorithm for out-of-core matrix transposition," in IEEE Transactions on Computers, vol. 51, no. 4, pp. 420-438, April 2002, doi: 10.1109/12.995452.
\bibitem{b20} Kandemir, M. \& Choudhary, A. \& Ramanujam, J. \& Arunachalam, Meenakshi. (2000). A unified framework for optimizing locality, parallelism, and communication in out-of-core computations. Parallel and Distributed Systems, IEEE Transactions on. 11. 648 - 668. 10.1109/71.877759. 
\bibitem{b21} I-Jui Sung, Juan G´omez-Luna, Josè Marìa Gonzàlez-Linares
Nicolàs Gui, Wen-Mei W. Hwu. In-Place Transposition of Rectangular Matrices on Accelerators
\bibitem{b22} Post: An Efficient Matrix Transpose in CUDA C/C++. Feb 18, 2013 By Mark Harris, https://developer.nvidia.com/blog/efficient-matrix-transpose-cuda-cc/
\bibitem{b23} Hanounik, Bedros. (2000). Diagonal Registers: Novel Vector Register File Design for High Performance and Multimedia Computing. 
\end{thebibliography}
\vspace{12pt}
